Impact of Demographics and Skills of the Observer on Visual Statistical Inference

### 2013-05-30 sociology_paper.tex
conclusion added

The test statistic used in visual inference is a plot of the data. This plot is then placed on a layout of plot called lineup. The rest of the plots in a lineup are generated from the model specified by the null hypothesis. An observer is then asked to evaluate the lineup. If the observer can detect the observed plot in the lineup, the hypothesis is rejected. Figure \ref{fig:lineup_geno} displays a lineup of 20 plots where one of the plots is observed data. The rest of the 19 plots are generated from the assumption that there is no difference between two groups of data. For this, the group structure of the data is broken by random permutation of the groups. If the null hypothesis is true, the observed data should look similar to the rest of the plots in the lineup making it difficult to detect the observed plot. On the other hand if the null hypothesis is not true, observed data should be different from the rest of the plots in the lineup.

@@

For a lineup with fixed difficulty, the proportion correct responses may vary for different observers. If an observe evaluates multiple lineups, the proportion of correct evaluations by the observer could be used as a measure of performance by an observer. The another way of measuring the performance of the observer could be the time taken by an observer to evaluate a lineup. Since performance would also depend on the lineup difficulty it is essential to adjust for this before estimating the observer performance.

@@

The power can be estimated from the proportion of correct evaluations. Thus power could be used as a measure of performance of the observer.

@@

\section{Simulation Experiment and Methods}\label{sec:exp_design}
The performance of the observer while evaluating a lineup would depend on the factors described in Section \ref{sec:facotr_performance}. In this section we present the experimental design to control some  of them so that learning effect of observer and location effect of a plot in the lineup can be estimated.

@@

@@

\section{Results}\label{sec:result_socio}
\subsection{Overview of the Data}
\subsection{Diversity of the Experimental Subjects} Amazon Mechanical Turk \cite{turk} web site is a good source for recruiting people to evaluation lineups. It is a source of diverse workers who can give us feedback on visual inspection.  Figure \ref{fig:turker_location} shows the location from where we received data. Notice that we received data from around the world and we observed diversity of location in the collected data. We notice diversity in not only location of the observers but also their gender, age groups and education levels. Table \ref{tbl:demographics} shows that number of male and female participants are almost same.

@@

%covariates attempt number and $p$-value of the lineup. The $p$-value of the lineup measures some sort of difficulty and including this should adjust for the difficulty level of the lineup while estimating the probability of correct responses in each attempt. Figure \ref{fig:learning_trend} shows the fitted subject wise learning pattern as well as overall learning pattern for each of the experiments. The parameter estimates of the model are shown in Table \ref{tbl:model_par}.

@@

\appendix

\begin{table}[hbtp]
\caption{Demographic information of the subjects participated the MTurk experiments.}
\centering
\begin{tabular}{rlrrr}
  \hline
 & levels & subsjects & avg\_time & response \\ 
  \hline
Gender & Male & 1348 & 48.51 & 13493 \\ 
   & Female & 991 & 43.75 & 10564 \\ 
\hline
  Education & High school or less & 193 & 37.21 & 2241 \\ 
   & Some under graduate courses & 418 & 42.84 & 4070 \\ 
   & Under graduate degree & 584 & 44.29 & 5775 \\ 
   & Some graduate courses & 245 & 43.43 & 2460 \\ 
   & Graduate degree & 902 & 52.18 & 9511 \\ 
\hline
  Age & 18-25 & 740 & 42.97 & 7311 \\ 
   & 26-30 & 547 & 46.27 & 5585 \\ 
   & 31-35 & 376 & 44.27 & 3923 \\ 
   & 36-40 & 257 & 55.03 & 2714 \\ 
   & 41-45 & 141 & 43.90 & 1519 \\ 
   & 46-50 &  95 & 49.29 & 1003 \\ 
   & 51-55 &  83 & 48.67 & 867 \\ 
   & 56-60 &  64 & 59.73 & 678 \\ 
   & above 60 &  38 & 48.67 & 457 \\ 
   \hline
\end{tabular}\label{tbl:demographics}
\end{table}

>>>>

The test statistic used in visual inference is a plot of the data. This plot is then placed on a layout of plot called lineup. The rest of the plots in a lineup are generated from the model specified by the null hypothesis. An observer is then asked to evaluate the lineup. If the observer can detect the actual data plot in the lineup, the hypothesis is rejected. Figure \ref{fig:lineup_geno} displays a lineup of 20 plots where one of the plots is observed data. The rest of the 19 plots are generated from the assumption that there is no difference between two groups of data. For this, the group structure of the data is broken by random permutation of the groups. If the null hypothesis is true, the observed data should look similar to the rest of the plots in the lineup making it difficult to detect the observed plot. On the other hand if the null hypothesis is not true, observed data should be different from the rest of the plots in the lineup.

@@

For a lineup with fixed difficulty, the proportion correct responses may vary for different observers. If an observer evaluates multiple lineups, the proportion of correct evaluations by the observer could be used as a measure of performance by an observer. The another way of measuring the performance of the observer could be the time taken by an observer to evaluate a lineup. Since performance would also depend on the lineup difficulty it is essential to adjust for lineup difficulty before estimating the performance of the observer.

@@

The visual test statistic is chosen so that it displays a specific patter in case the null hypothesis not true. Thus the most important factor that help an observer to correctly evaluate a lineup is the presence of any detectable signal in the data. On the other hand, if the null hypothesis is not true, visual test statistic should not display any distinguishable pattern. In fact, some of the null plots may appear to be the most different plot in a lineup influencing the observer to chose a plot different than actual data plot. This is an elegant feature of the lineup. It force the observer to chose a wrong plot when the null hypothesis should not be rejected.

@@

\section{Simulation Experiment and Methods}\label{sec:exp_design}
The performance of the observer while evaluating a lineup would depend on the factors described in Section \ref{sec:factor_performance}. In this section we present the experimental design to control some  of them so that learning effect of observer and location effect of a plot in the lineup can be estimated.

@@

As per the design of the experiment for learning trend we have two different response variables. One is observer feedback on a lineup which is a binary variable.  The other variable is the time taken for each evaluation which is a continuous variable. Suppose random variable $Y_{\ell, i}$ denotes the response from subject $ i=1,2, ..., K$ on lineup $\ell=1,2, ..., L$. We model the mean $\mu_{\ell, i} = E(Y_{\ell, i})$ with a generalized liner mixed effect model

\begin{equation} \label{eqn:trend}
g( \mu_{\ell, i} )= W_{\ell i} \delta +  Z_{\ell i}  \tau_{\ell i},  
\end{equation}

\noindent where $g(.)$ denotes the link function. $W$ is a design matrix of covariates corresponding to specifics of lineup $\ell$ and subject $i$, and $\delta$ is the vector of corresponding parameters. Covariates could include  demographic information of individuals, such as age, gender, education level etc.,  as well as the attempt made by the observer. $Z_{\ell i}$,  $1 \le i \le K$, $1 \le \ell \le L$,  is a design matrix corresponding to random effects specific to individual $i$ and lineup $\ell$; and 
$\tau$  is a vector of independent normally distributed random variables $\tau_{\ell i}$ with  variance matrix $\sigma_\tau I_{KL \times KL}$. $\tau$ will usually include a component incorporating an individual's ability or skill to evaluate lineups.

The inverse link function, $g^{-1}(.)$, from  Equation \ref{eqn:trend} leads to the estimate of the subject and the lineup specific probability of successful evaluation by a single observer as 
\begin{equation} \label{eqn:trend_power}
\hat p_{\ell i} =  g^{-1}(W_{\ell i} \hat {\delta} +  Z_{\ell i}  \hat {\tau}_{\ell i}).
\end{equation}

For $Y_{l,i}$ being binary response we use {\it logit} link function $g(\pi)=\log(\pi) - \log(1-\pi); 0 \le \pi \le 1$ and $\hat p_{\ell i}$ gives the estimate of proportion correct by subject $i$ with lineup $\ell$. For continuous response  $Y_{l,i}$ we use gamma link function and $\hat p_{\ell i}$ gives the estimate of mean time taken by subject $i$ for lineup $\ell$.

@@

\section{Results}\label{sec:result_socio}

\subsection{Overview of the Data}

\begin{table}[hbtp]
\caption{Demographic information of the subjects participated the MTurk experiments.}
\centering
\begin{tabular}{rlrrr}
  \hline
 & levels & subsjects & avg\_time & response \\ 
  \hline
Gender & Male & 1348 & 48.51 & 13493 \\ 
   & Female & 991 & 43.75 & 10564 \\ 
\hline
  Education & High school or less & 193 & 37.21 & 2241 \\ 
   & Some under graduate courses & 418 & 42.84 & 4070 \\ 
   & Under graduate degree & 584 & 44.29 & 5775 \\ 
   & Some graduate courses & 245 & 43.43 & 2460 \\ 
   & Graduate degree & 902 & 52.18 & 9511 \\ 
\hline
  Age & 18-25 & 740 & 42.97 & 7311 \\ 
   & 26-30 & 547 & 46.27 & 5585 \\ 
   & 31-35 & 376 & 44.27 & 3923 \\ 
   & 36-40 & 257 & 55.03 & 2714 \\ 
   & 41-45 & 141 & 43.90 & 1519 \\ 
   & 46-50 &  95 & 49.29 & 1003 \\ 
   & 51-55 &  83 & 48.67 & 867 \\ 
   & 56-60 &  64 & 59.73 & 678 \\ 
   & above 60 &  38 & 48.67 & 457 \\ 
\hline   
  Country  & United States & 1087 & 39.64 & 10769 \\  
  & India & 980 & 52.63 & 10227 \\ 
  & Rest of the world & 254 & 46.86 & 2819 \\

   \hline
\end{tabular}\label{tbl:demographics}
\end{table}

\subsection{Diversity of the Experimental Subjects} Amazon Mechanical Turk \cite{turk} web site is a good source for recruiting people to evaluation lineups. It is a source of diverse workers who can give us feedback on visual inspection.  Figure \ref{fig:turker_location} shows the location from where we received data. Notice that we received data from around the world and we observed diversity of location in the collected data. We notice diversity in not only location of the observers but also their gender, age groups and education levels. Table \ref{tbl:demographics} shows that number of male and female participants are almost same.

@@

@@

With controlled simulation experiments, it is observed that the performance of human observers does not increase through successive attempts while evaluating multiple lineups. Even though the performance does not significantly improve, their skill in evaluating the lineup in shorter time gets improved over successive evaluations. When an observer evaluates multiple lineups, the earlier evaluations take significantly longer than the later evaluations. This result is very important as it suggests that the power estimated for visual inference using human subject experiment is robust and may not change if those participants are allowed to give feedback again. This investigation suggest that the skilled person may only do it faster.

The another simulation experiment reveals that the effect of location of actual data plot in the lineup is not significant. This is important as the visual statistical inference procedure suggest that the data plot be placed at random anywhere in the lineup.
This paper suggest that any random place in a lineup is as good as other places in the lineup.

More experiment may be needed to make a concrete decision about the difference in power with skilled observer and a non-skilled observer in terms of statistical graphics. The numerous pilot studies done with the more trained participants suggest that power of visual statical inference is higher with skilled observer.

This investigation allowed each observer to evaluate 10 lineups assuming that it would not cause fatigue or disinterest of the observers towards the experiment based on the pilot studies. The future research may involve doing more experiments to check if fewer or more than 10 lineups have any significant impact on performance of the observer. Also a lineup with fewer or larger than 10 plots may show different results. If the size of the lineup is much higher than 20, it is intuitive that there may be location effect of the data plot in the lineup.

\section{Some figures and tables (for review)}

\begin{figure}[htbp] 
   \centering
   \includegraphics[width=6in]{turker_location_experiment.pdf} 
   \caption{World maps showing where the participants are coming from for all the 10 experiments.}
   \label{fig:turker_location_experiment}
\end{figure}
